Relationship between solar wind, $D_{st}$, and plasmasphere trough mass density on one-hour time scales

### 2016-03-18 paper.tex
Compiling paper

The plasma trough is a transition region between the high density and low temperature plasmasphere and the lower density and higher energy magnetosphere \citep{Mayr1968ModelMagnetosphereTemperature}.  Inside of the plasmapause, the density is primarily due to ionospheric photoionization; outside of the plasmapuase, the density has contributions from the co-rotating plasmasphere and convecting magnetosphere plasmas.

@@

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 (obervations were made when CRRESS was in an elliptical, near equatorial orbit and primarily in the plasma trough region). They found that the average ion mass, $M=\rho/n_e$, had some correlation with geomagnetic activity, with more negative hourly $D_{st}$ values corresponding to higher $M$ and 1.5- and 3-day averages of $K_p$ corresponding to higher $M$ (3-hour averages of $K_p$had little visual correlation with $M$). The mass density estimates were obtained from CRRES observations in the MLT range of 12:00 and 18:00 with most observations having L between 5 and 7 $R_E$.

\cite{Denton2006} found a weak relationship between $D_{st}$ and and $K_p$ and the field line distribution of $\rho$ using observations of the first three torodial Alfven harmonic fequencies for $L$ in the range of 6-8$R_E$.  A more pronounced peak in the distribution along a filed line near the equator was fround for lower $D_{st}$ and larger $K_p$.  The equatorial value of $\rho$ appeared identical for the bins of $D_{st}$ (-142 to -31 nT and -31 to 37 nT) and $K_p$ (1.5 to 3.4 and 3.4 to 5.9) considered .

@@

\cite{Takahashi2010} noted that downward spikes in the $D_{st}$ index coincided with with significant changes in $\rho_{eq}$ for $L$ near 6.8~$R_E$. For five storms, two had $\rho_{eq}$ spikes after the $D_{st}$ drop, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, an enhancment in $\rho_{eq}$ appeared on the same day as minimum $D_{st}$.

\cite{Yao2008} studied the relationship between $D_{st}$ and ion number density for low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-sheels (2-14) had a strong correlation (0.88).  Although the correlation in the plasma trough was not calculated, the data presented show that the variations in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

@@

Issue (1) is addressed by considering in detail the number of observations that are used in epoch averages around the time of two types of events, decreases in $D_{st}$ below a threshold, and increases in $\rho_{eq}$ above a threshold.  Issue (2) is addressed using various methods to remove or isolate the F10.7 depenences.  Issue (3) is addressed by considering epochs and linear correlations independent of the direction of the IMF along with separation of the epoch averages based on the northrward and southward IMF.

In addition, we consider the influence of time averaging by comparing results using daily means and meadians and hourly means and medians of solar wind and magentospheric parameters.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset of \cite{Denton}, which was used in \cite{Takahashi2010}; data are available from 1980 through 1991.  All other parameters used are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements as described in \cite{Takahashi2010}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hour window was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available in the \cite{Denton} dataset from multiple GOES satellites at the same time, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

@@

Figure~\ref{ccplot} shows the long-term trends of GOES 6 based $log(\rho_{eq})$ and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows using all non-fill hourly values.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot}.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0 and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites are 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

@@

Two types of events are considered. The first is a drop in the $D_{st}$ index \vnote{Cite Kyoto} below the threshold of -50~nT. These types of events are considered in order to determine if there is a well-defined $\rho_{eq}$ dependence on geomagnetic activity/storms as indicated by $D_{st}$ index.  It is expected that the plasmasphere will exhibit two responses near these threshold crossings.  First is the movement of the plasmasphere earthward and the second is a possible increase in density due to magnetospheric processes \vnote{cite}.

@@

Our first analysis uses data only from the time interval 1989-1991.  In this interval, there were 329 $D_{st}$ events and 70 $\rho_{eq}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians accross events.)  Error bars for each parameter were computed using the standard deviation of values used in computing the median dividied by the square root of the number of values used.

In the upper panel of \ref{DailyAveragedDstEvents}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events corrspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is on the order of $\sim2$~amu/cm$^3$ instead of $\sim10$~amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for $D_{st}$, the number of measurements used in computing the medians is much larger as we use all available $D_{st}$ measurements instead of restricting to only values where a non-fill $\rho_{eq}$ value existed.

Both of the results in \ref{DailyAveragedDstEvents} indicate that there is not a statictically significant change in $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.

Because the dynamics of the magnetopshere take place on a time scale of less than one day, we have also considered hourly averages as shown in the top panel of \ref{HourlyAveragedDstEvents}.  A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours.  The error bars were computed in the same way as \ref{DailyAveragedDstEvents}.  In this plot, the overall trend is similar to that found in the bottom panel of \ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slighly elevated at the time of the threhshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small and in general not statistically significant.

@@

An alterntive type of event can be defined that corresponds to large increases in $\rho\ $.  Figure~\ref{HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\req\ $ above $20$~amu/cm$^2$ defines the start of the event.  Here we see the expected result that such events tend to be associated with positive $B_z^{IMF}$.  To determine if the sign of $B_z^{IMF}$ influences the $\rho\ $ epoch time series, we separated the events by the average of $B_z^{IMF}$ in a time window of $4$~hours centered on the event start time.  The epoch curves for $B_z^{IMF}\geq=0$ (257) events and $B_z^{IMF}\ge=0$ (139) have no statistically signifiant difference.

@@

\begin{figure}[htp!]
	\centering
	\includegraphics[scale=0.40]{paperfigures/stormavs-mass-gt20-GOES6.eps}
\caption{$\req\ > 20$~ amu/cm^3 events from GOES 6 using hourly medians. As described in the text, there is no statistically significant dependence of the $\req\ $ plot on $B_z$ in the time interval of $\pm 2$~hours around the event.}
\label{HourlyAveragedRhoEvents}
\end{figure}

>>>>

The plasma trough is a transition region between the high density and low temperature plasmasphere and the lower density and higher energy magnetosphere \citep{Mayr1968ModelMagnetosphereTemperature}.  Inside of the plasmapause, the density is primarily due to ionospheric photoionization; outside of the plasmapause, the density has contributions from the co-rotating plasmasphere and convecting magnetosphere plasmas.

@@

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 (observations were made when CRRESS was in an elliptical, near equatorial orbit and primarily in the plasma trough region). They found that the average ion mass, $M=\rho/n_e$, had some correlation with geomagnetic activity, with more negative hourly $D_{st}$ values corresponding to higher $M$ and 1.5- and 3-day averages of $K_p$ corresponding to higher $M$ (3-hour averages of $K_p$had little visual correlation with $M$). The mass density estimates were obtained from CRRES observations in the MLT range of 12:00 and 18:00 with most observations having L between 5 and 7 $R_E$.

\cite{Denton2006} found a weak relationship between $D_{st}$ and and $K_p$ and the field line distribution of $\rho$ using observations of the first three toroidal Alfvén harmonic frequencies for $L$ in the range of 6-8$R_E$.  A more pronounced peak in the distribution along a filed line near the equator was found for lower $D_{st}$ and larger $K_p$.  The equatorial value of $\rho$ appeared identical for the bins of $D_{st}$ (-142 to -31 nT and -31 to 37 nT) and $K_p$ (1.5 to 3.4 and 3.4 to 5.9) considered .

@@

\cite{Takahashi2010} noted that downward spikes in the $D_{st}$ index coincided with with significant changes in $\rho_{eq}$ for $L$ near 6.8~$R_E$. For five storms, two had $\rho_{eq}$ spikes after the $D_{st}$ drop, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, an enhancement in $\rho_{eq}$ appeared on the same day as minimum $D_{st}$.

\cite{Yao2008} studied the relationship between $D_{st}$ and ion number density for low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-shells (2-14) had a strong correlation (0.88).  Although the correlation in the plasma trough was not calculated, the data presented show that the variations in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

@@

Issue (1) is addressed by considering in detail the number of observations that are used in epoch averages around the time of two types of events, decreases in $D_{st}$ below a threshold, and increases in $\rho_{eq}$ above a threshold.  Issue (2) is addressed using various methods to remove or isolate the F10.7 dependences.  Issue (3) is addressed by considering epochs and linear correlations independent of the direction of the IMF along with separation of the epoch averages based on the northward and southward IMF.

In addition, we consider the influence of time averaging by comparing results using daily means and medians and hourly means and medians of solar wind and magnetospheric parameters.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset of \cite{Denton}, which was used in \cite{Takahashi2010}; data are available from 1980 through 1991.  All other parameters used are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic toroidal frequency of magnetic field measurements as described in \cite{Takahashi2010}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hour window was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available in the \cite{Denton} dataset from multiple GOES satellites at the same time, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

@@

Figure~\ref{ccplot} shows the long-term trends of GOES 6 based $log(\rho_{eq})$ and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows using all non-fill hourly values.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot}.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0$ and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites are 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

@@

Two types of events are considered. The first is a drop in the $D_{st}$ index \vnote{Cite Kyoto} below the threshold of -50~nT. These types of events are considered in order to determine if there is a well-defined $\rho_{eq}$ dependence on geomagnetic activity/storms as indicated by $D_{st}$ index.  It is expected that the plasmasphere will exhibit two responses near these threshold crossings.  First is the movement of the plasmasphere earthward and the second is a possible increase in density due to magnetospheric processes \inote{cite}.

@@

Our first analysis uses data only from the time interval 1989-1991.  In this interval, there were 329 $D_{st}$ events and 70 $\rho_{eq}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians across events.)  Error bars for each parameter were computed using the standard deviation of values used in computing the median divided by the square root of the number of values used.

In the upper panel of \ref{DailyAveragedDstEvents}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is on the order of $\sim2$~amu/cm$^3$ instead of $\sim10$~amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for $D_{st}$, the number of measurements used in computing the medians is much larger as we use all available $D_{st}$ measurements instead of restricting to only values where a non-fill $\rho_{eq}$ value existed.

Both of the results in \ref{DailyAveragedDstEvents} indicate that there is not a statistically significant change in $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.

Because the dynamics of the magnetosphere take place on a time scale of less than one day, we have also considered hourly averages as shown in the top panel of \ref{HourlyAveragedDstEvents}.  A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours.  The error bars were computed in the same way as \ref{DailyAveragedDstEvents}.  In this plot, the overall trend is similar to that found in the bottom panel of \ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slightly elevated at the time of the threshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small and in general not statistically significant.

@@

An alternative type of event can be defined that corresponds to large increases in $\rho\ $.  Figure~\ref{HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\req\ $ above $20$~amu/cm$^2$ defines the start of the event.  Here we see the expected result that such events tend to be associated with positive $B_z^{IMF}$.  To determine if the sign of $B_z^{IMF}$ influences the $\rho\ $ epoch time series, we separated the events by the average of $B_z^{IMF}$ in a time window of $4$~hours centered on the event start time.  The epoch curves for $B_z^{IMF}\geq=0$ (257) events and $B_z^{IMF}\ge=0$ (139) have no statistically significant difference.

@@

\begin{figure}[htp!]
	\centering
	\includegraphics[scale=0.40]{paperfigures/stormavs-mass-gt20-GOES6.eps}
\caption{$\req\ > 20$~ amu/cm$^3$ events from GOES 6 using hourly medians. As described in the text, there is no statistically significant dependence of the \req\  plot on $B_z$ in the time interval of $\pm 2$~hours around the event.}
\label{HourlyAveragedRhoEvents}
\end{figure}
